% Indicate the main file. Must go at the beginning of the file.
% !TEX root = ../main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 01_introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}\todo{Its a starting point, needs to be extended and improved.}
\label{introduction}

The ongoing loss of biodiversity is among the most urgent environmental issues globally \autocite{brondizioGlobalAssessmentReport2019, cardinaleBiodiversityLossIts2012}.
Small mammals, despite their ecological importance, often receive limited attention in conservation research compared to birds and butterflies \autocite{grafWildlifeCampusKleineSaeugetiere2022}.
This disparity persists although small mammals significantly contribute to ecosystem functions such as seed dispersal and soil aeration.
In Switzerland, of the approximately 30 native small mammal species, several are endangered and require targeted conservation strategies \autocite{bafuListeNationalPrioritaren2019}.
In response to these challenges, the Wildlife@Campus project was initiated to establish a long-term and locally anchored monitoring program for small mammals in the urban context of the \ac{ZHAW} Campus Wädenswil.
By combining citizen science with scientific research, the project aims to improve the detection and understanding of small mammal populations while also raising public awareness and contributing to species conservation efforts \autocite{grafWildlifeCampusKleineSaeugetiere2022}.

\subsection{Background}
Monitoring small mammal populations traditionally involves labor-intensive methods like live trapping, which not only consume significant time and resources but also pose risks to animal welfare \autocite{grafWildlifeCampusKleineSaeugetiere2022}.
Indirect, more efficient survey methodes are a very much needed alternative to monitor less invasive and to generate more data for less effort.
A proven to be effective method is the use of footprint tunnels showing a very low non detection rate \autocite{yarnellUsingOccupancyAnalysis2014}.
A further non-invasive method with increasing relevance is the use of \ac{eDNA} for species detection with high reliability \autocite{thomsenEnvironmentalDNAEmerging2015}.
The method in focus for this study utilizes camera traps for species detection --- an approach that has seen rapid and widespread adoption in ecological and conservation research \autocite{delisleNextGenerationCameraTrapping2021}.
Based on findings from the Wildlife@Campus project, a combined approach using camera traps for continuous monitoring and \acs{eDNA}-based methods for species-level identification appears to be a promising and scalable solution for small mammal detection in urban environments \autocite{grafWildlifeCampusKleineSaeugetiere2022}.
As part of the Wildlife@Campus project, initial efforts were made to explore the collection of non-invasive genetic material such as hair and fecal samples.
While these genetic methods were not fully implemented during the pilot phase, they were recognized as essential for identifying cryptic or morphologically similar species and are considered a key component of future monitoring strategies.
Based on the design of the \enquote{Mostela} camera trap box, \textcite{aegerterMonitoringKleinmustelidenSchlaefern2019} developed the \enquote{MammaliaBox}, adapting it to local conditions for monitoring small mustelids and dormice in Switzerland.
In addition to improving image quality, he conducted extensive field testing and produced a detailed guideline for standardized and effective deployment.
During the Wildlife@Campus project, images were generated exclusively using \enquote{MammaliaBoxes} at multiple locations on the ZHAW campus. 
These camera trap boxes produced a substantial volume of image material, capturing a wide range of small mammal activity.
Significant effort has already been invested within the Wildlife@Campus project to manually label a substantial portion of the image dataset and to explore automated approaches to species detection.
In fact, a preliminary solution for training a \ac{DL} model was developed and tested, and a usable labeled dataset already exists \autocite{ratnaweeraWildlifeCampusProgressReports2021}.
However, this work remained largely undocumented and was not brought to full completion. 
In the context of this bachelor's thesis, this dataset will be revisited to explore possibilities to leverage \ac{DL} techniques for species detection in small mammel images.

\subsection{Problem Statement}

The aim of this thesis is to develop a reproducible and effective workflow for the automated classification of small mammals in camera trap images. To achieve this, the following tasks are addressed:

\begin{itemize}
    \item \textbf{Dataset Preparation:} Review and analyze the existing labeled dataset, including understanding the data structure and labeling conventions.
    
    \item \textbf{Data Processing:} Develop a processing pipeline that includes identifying the \ac{ROI} in each image, selecting appropriate training samples, and preparing the data for \ac{DL} model training.
    
    \item \textbf{Model Selection:} Research and select suitable \ac{DL} architectures for the classification task, considering both performance and computational efficiency.
    
    \item \textbf{Model Training:} Train the selected models on the prepared dataset using appropriate optimization strategies.
    
    \item \textbf{Model Evaluation:} Assess and compare model performance using standard evaluation metrics, and conduct a detailed analysis of the most promising architecture.
\end{itemize}


\subsection{Related Work}

% Deap Learning for image classification
Image classification utilizing \ac{DL} has become a cornerstone of modern computer vision, with significant advancements in recent years.
One of the earliest successfull \ac{CNN} architectures was LeNet-5, introduced by \textcite{lecunGradientbasedLearningApplied1998} --- it was designed for handwritten digit recognition and laid the foundation for subsequent developments in \ac{DL} for image classification.
A major breakthrough came with the introduction of AlexNet by \textcite{krizhevskyImageNetClassificationDeep2012}, which demonstrated the power of deep convolutional networks in large-scale image classification tasks.
This architecture won the ImageNet Large Scale Visual Recognition Challenge (ILSVRC) in 2012 and significantly outperformed previous methods, showcasing the potential of \ac{DL} in computer vision.
Following AlexNet, several other influential architectures emerged, including VGGNet \autocite{simonyanVeryDeepConvolutional2015}, GoogLeNet \autocite{szegedyGoingDeeperConvolutions2015}, and ResNet \autocite{heDeepResidualLearning2016}.
These architectures introduced various innovations, such as deeper networks, inception modules, and residual connections, which further improved classification accuracy and model efficiency.
The development of these architectures has been complemented by the creation of large-scale datasets like ImageNet \autocite{dengImageNetLargescaleHierarchical2009}, which provided a rich source of labeled images for training and benchmarking \ac{DL} models.
The are of modern \ac{DL} for image classification has continued to evolve, with the introduction of transformer-based architectures like \ac{ViT} \autocite{dosovitskiyImageWorth16x162021} and Swin Transformers \autocite{liuSwinTransformerHierarchical2021}, which have shown promising results in various image classification tasks.


% Deep Learning for wildlife monitoring
The success of \ac{DL} in general image classification has encouraged its adoption in wildlife monitoring, where camera trap images pose unique challenges due to varying lighting, occlusion, species similarity, and low data quality.

% use of megadetector

There is a growing number of \ac{AI} platforms and tools designed to facilitate the application of \ac{DL} in wildlife monitoring.
\textcite{velezChoosingAppropriatePlatform2022} provide a comprehensive overview of various platforms, including \ac{MD}, \ac{WI}, \ac{MLWIC2}, and Conservation \ac{AI}.
\ac{WI} and \ac{MLWIC2} demonstrated low recall—many animals present in images were missed—but high precision for some species.
In contrast, \ac{MD} performed reliably for broad classifications, such as distinguishing \enquote{animal} from \enquote{blank}.
The authors conclude that while fully automated species classification is not yet reliable, \ac{DL} tools are valuable for filtering blank images and supporting semi-automated workflows.
\ac{MD}, originally developed by Microsoft, has since become one of the most widely adopted tools for wildlife image filtering and is actively maintained as a standalone detection model \autocite{morrisEfficientPipelineCamera2025}.
Building on this foundation, a highly regarded research team supported by Microsoft is now developing the PyTorch Wildlife project \autocite{hernandezPytorchWildlifeCollaborativeDeep2024}.
This framework incorporates \ac{MD} as a core component and extends its functionality into a full-featured \ac{DL} platform tailored to wildlife monitoring.
A study applying \ac{MD} for image preprocessing is presented by \textcite{schneiderRecognitionEuropeanMammals2024}, who demonstrate its effectiveness in filtering blank images and identifying animals in camera trap datasets.
A particularly notable contribution of their work is the taxonomic classification approach, enabling hierarchical predictions beyond species level to genus, family, or order.


% studies focussing on small mammals


% studies focussing on sequential detection

There are more complex approaches leveraging the often sequential nature of camera trap data, to improve detection and classification accuracy.
As an example, \textcite{zotinANIMALDETECTIONUSING2019} utilizes sequences for a none \ac{DL} approach to detect animals in camera trap images taken under complex shooting conditions.
Another study by \textcite{muhammadTemporalSwinFPNNetNovel2024} introduces a sequence-aware and metadata-driven approach to camera trap image classification by adapting the Swin Transformer architecture (Swin-FPN Net) to process image sequences rather than individual frames. 
By incorporating temporal dynamics and leveraging metadata such as timestamps, their method enhances classification accuracy and reduces processing time, enabling a more context-sensitive and efficient analysis of wildlife imagery.
